Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
fixed figure with Eliot
Add a reference for Monticello.  Delete middleabstract para p[er Elisa's suggestion

\egb{I think you can skip the next paragraph from the abstract. If the point is that you need to co-evolve both execution models once the VM changes, you can do that with 1 line in the previous paragraph. Now it may look like you will discuss co-evolution issues, but I guess you do not}

As the VM has evolved, including the addition of a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built an open-source Smalltalk \egb{dialect?} and VM, Squeak \cite{SqueakByExample}, using the code from \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
\egb{Changed again this paragraph after reading section 2. Maybe we should also define here what simulation and production is (since the abstract is not considered to be part of the main body of the paper)}
Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.  
Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM.

At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
\item and external C code to support running the VM on popular platforms.
\end{itemize}

@@

>>>>

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built Squeak \cite{BackToTheFuture}, an open-source Smalltalk dialect, Squeak \cite{SqueakByExample}, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.  Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.

When the Squeak VM was released it consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
\item external C code and makefiles to support running the VM on popular platforms.
\end{itemize}

@@

In this paper we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
